---
title: "NMDS Ordination of Macroinvertebrate Community Composition"
subtitle: "Bray-Curtis Dissimilarity | Abundance-Weighted | Gradient & Season Comparison"
author: "Methods in Stream Ecology"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 6,
  dpi        = 300
)
```

------------------------------------------------------------------------

## Overview

This tutorial walks through a **Nonmetric Multidimensional Scaling (NMDS)** ordination of macroinvertebrate abundance data to compare community composition across stream gradient types (Riffle vs. Boatable) and seasons (Spring vs. Fall).

NMDS is a non-parametric ordination technique well-suited for ecological community data because:

-   It ranks dissimilarities rather than assuming multivariate normality
-   It works with any distance/dissimilarity metric (we use **Bray-Curtis** here)
-   It is robust to the many zeros common in species-abundance matrices
-   It provides a low-dimensional visual summary of community variation (McCune & Grace 2002)

**Ecological question:** Does macroinvertebrate community composition differ between **stream gradient types** (Riffle vs. Boatable) and/or **seasons** (Spring vs. Fall)?

------------------------------------------------------------------------

## Data Structure

This workflow uses three input files that must be in your `data/` folder:

| File | Contents | Key columns |
|----|----|----|
| `stationBenthicsTESTSITE.csv` | Raw macroinvertebrate counts (long format) | `StationID`, `BenSampID`, `FinalID`, `Individuals` |
| `stationInfoBenSampsTESTSITE.csv` | Sample metadata | `BenSampID`, `Gradient`, `Season` |
| `masterTaxaGenus.csv` | Taxonomic reference and traits | `FinalID`, `Order`, `Family`, `FFG`, `TolVal` |

These files are joined by `BenSampID` (unique sample identifier) and `FinalID` (taxon name).

------------------------------------------------------------------------

## Step 1: Load Required Packages

```{r packages}
# Install packages if needed (uncomment lines below)
# install.packages(c("dplyr", "tidyr", "ggplot2", "vegan", "ggrepel", "indicspecies"))

library(dplyr)        # data wrangling
library(tidyr)        # pivoting to wide format
library(ggplot2)      # visualization
library(vegan)        # NMDS, PERMANOVA, Bray-Curtis (Oksanen et al. 2022)
library(ggrepel)      # non-overlapping text labels
library(indicspecies) # indicator species analysis (De Caceres & Legendre 2009)
```

**Key packages and citations:**

| Package | Purpose | Citation |
|----|----|----|
| `vegan` | NMDS, PERMANOVA, Bray-Curtis dissimilarity | Oksanen et al. (2022) |
| `indicspecies` | Indicator species analysis (IndVal) | De Caceres & Legendre (2009) |
| `ggplot2` | Visualization | Wickham (2016) |

------------------------------------------------------------------------

## Step 2: Load Data

```{r load-data}
benthics <- read.csv("~/Dropbox/VT/MethodsChapter/NMDS/R Module for Github/Data/stationBenthicsTESTSITE.csv")

site_info <- read.csv("~/Dropbox/VT/MethodsChapter/NMDS/R Module for Github/Data/stationInfoBenSampsTESTSITE.csv")

master_taxa <- read.csv("~/Dropbox/VT/MethodsChapter/NMDS/R Module for Github/Data/masterTaxaGenus.csv")

# Inspect structure
glimpse(benthics)
glimpse(site_info)
```

------------------------------------------------------------------------

## Step 3: Join Metadata to Benthic Counts

Each sample (`BenSampID`) needs its gradient type and season joined from the site info table. We also exclude samples collected outside the standard spring/fall sampling window.

```{r join-metadata}
# Select only the metadata columns we need
meta_slim <- site_info %>%
  select(StationID, BenSampID, RepNum, Gradient, Season) %>%
  filter(Season %in% c("Spring", "Fall"))   # drop "Outside Sample Window"
meta_slim

# Join metadata to benthic counts
benthics_meta <- benthics %>%
  inner_join(meta_slim, by = c("StationID", "BenSampID", "RepNum"))

# Check sample representation by gradient and season
benthics_meta %>% count(Gradient, Season)
```

> **Note:** Samples labeled `"Outside Sample Window"` are excluded because they fall outside the standard spring/fall sampling periods. Adjust this filter if your question requires them.

------------------------------------------------------------------------

## Step 4: Filter and Clean Counts

```{r filter-counts}
benthics_clean <- benthics_meta %>%
  filter(
    Excluded.Taxa == 0,   # remove flagged/excluded taxa
    Individuals   > 0,   # remove zero-count rows
    !is.na(FinalID)       # remove unidentified rows
  )

cat("Samples retained:", n_distinct(benthics_clean$BenSampID), "\n")
cat("Taxa retained:   ", n_distinct(benthics_clean$FinalID),   "\n")
```

------------------------------------------------------------------------

## Step 5: Build the Community Matrix

NMDS requires a **wide-format community matrix** where:

-   **Rows** = sampling units (`BenSampID`)
-   **Columns** = taxa (`FinalID`)
-   **Values** = individual counts, with `0` for taxa absent from a sample

```{r community-matrix}
# Aggregate counts by sample x taxon (sums across reps if multiple exist)
benthics_agg <- benthics_clean %>%
  group_by(StationID, BenSampID, Gradient, Season, FinalID) %>%
  summarise(Individuals = sum(Individuals, na.rm = TRUE), .groups = "drop")

# Pivot to wide format — one row per BenSampID, one column per taxon
comm_wide <- benthics_agg %>%
  select(BenSampID, FinalID, Individuals) %>%
  pivot_wider(
    names_from  = FinalID,
    values_from = Individuals,
    values_fill = 0
  )

# Separate metadata from the community matrix
meta <- benthics_agg %>%
  select(BenSampID, StationID, Gradient, Season) %>%
  distinct()

# Align row order between meta and comm_wide
meta <- meta[match(comm_wide$BenSampID, meta$BenSampID), ]

comm_matrix <- comm_wide %>%
  select(-BenSampID) %>%
  as.matrix()

rownames(comm_matrix) <- comm_wide$BenSampID

# Verify dimensions
cat("Samples (rows):", nrow(comm_matrix), "\n")
cat("Taxa (columns):", ncol(comm_matrix), "\n")
```

------------------------------------------------------------------------

## Step 6: Transform and Standardize Abundance Data

Raw abundance data are right-skewed and dominated by a few highly abundant taxa (e.g., Chironomidae). Two sequential transformations address this before calculating dissimilarities:

1.  **log(x + 1) transformation** — compresses the range of dominant taxa while preserving zeros
2.  **Wisconsin double standardization** — divides by species maximum then by site total; standard practice for community NMDS (McCune & Grace 2002)

```{r transform}
# Step 6a: log(x + 1) transformation
comm_log <- log1p(comm_matrix)

# Step 6b: Wisconsin double standardization
comm_scaled <- wisconsin(comm_log)

cat("Transformation complete. Value range after scaling:\n")
cat("  Min:", round(min(comm_scaled), 4), "\n")
cat("  Max:", round(max(comm_scaled), 4), "\n")
```

> **Why these two steps?** The log transformation reduces the leverage of dominant taxa; Wisconsin standardization ensures that highly abundant families do not disproportionately drive the ordination geometry. Together they produce a more ecologically balanced distance matrix.

------------------------------------------------------------------------

## Step 7: Calculate Bray-Curtis Dissimilarity

Bray-Curtis dissimilarity (Bray & Curtis 1957) is the standard metric for ecological community data. It ranges from 0 (identical communities) to 1 (no shared taxa) and ignores joint absences — two samples that both lack a taxon are not considered more similar because of it.

```{r bray-curtis}
bc_dist <- vegdist(comm_scaled, method = "bray")

# Inspect the distribution of pairwise dissimilarities
summary(as.vector(bc_dist))
```

------------------------------------------------------------------------

## Step 8: Run the NMDS Ordination

NMDS iteratively repositions samples in 2D space to best preserve the rank order of pairwise dissimilarities. **Stress** measures goodness of fit:

| Stress    | Interpretation                   |
|-----------|----------------------------------|
| \< 0.05   | Excellent representation         |
| 0.05–0.10 | Good                             |
| 0.10–0.20 | Acceptable                       |
| \> 0.20   | Poor — consider k = 3 dimensions |

```{r nmds, results='hide'}
set.seed(123)  # for reproducibility

nmds <- metaMDS(
  comm_scaled,
  distance = "bray",
  k        = 2,       # 2-dimensional solution
  trymax   = 100,     # random starts to find stable solution
  trace    = FALSE
)
```

```{r nmds-stress}
cat("NMDS stress:", round(nmds$stress, 4), "\n")

# Shepard plot — visualizes how well 2D distances represent dissimilarities
stressplot(nmds)
```

------------------------------------------------------------------------

## Step 9: Extract NMDS Scores

```{r extract-scores}
# Extract site scores and join metadata
nmds_scores <- as.data.frame(scores(nmds, display = "sites")) %>%
  mutate(BenSampID = rownames(.)) %>%
  left_join(meta, by = "BenSampID")

# Centroids by gradient type
centroids_gradient <- nmds_scores %>%
  group_by(Gradient) %>%
  summarise(NMDS1 = mean(NMDS1), NMDS2 = mean(NMDS2), .groups = "drop")

# Centroids by season
centroids_season <- nmds_scores %>%
  group_by(Season) %>%
  summarise(NMDS1 = mean(NMDS1), NMDS2 = mean(NMDS2), .groups = "drop")

# Centroids by gradient x season combination
nmds_scores <- nmds_scores %>%
  mutate(GradSeason = paste(Gradient, Season, sep = "_"))

centroids_combo <- nmds_scores %>%
  group_by(GradSeason) %>%
  summarise(NMDS1 = mean(NMDS1), NMDS2 = mean(NMDS2), .groups = "drop")

head(nmds_scores)
```

------------------------------------------------------------------------

## Step 10: Plot NMDS — Gradient Type

```{r plot-gradient, fig.cap="NMDS ordination of macroinvertebrate community composition by stream gradient type. Points = individual samples (BenSampID); diamonds = group centroids; shaded ellipses = 95% confidence intervals. Bray-Curtis dissimilarity on log(x+1) Wisconsin-standardized abundance data."}

gradient_colors <- c(
  "Riffle"   = "#513dd1",
  "Boatable" = "#E15759"
)

nmds_plot_gradient <- ggplot(nmds_scores,
                             aes(x = NMDS1, y = NMDS2,
                                 color = Gradient, fill = Gradient)) +
  stat_ellipse(
    aes(group = Gradient),
    type = "t", level = 0.95,
    geom = "polygon", alpha = 0.10, color = NA
  ) +
  stat_ellipse(
    aes(group = Gradient),
    type = "t", level = 0.95, linewidth = 0.6
  ) +
  geom_point(alpha = 0.6, size = 2) +
  geom_point(
    data        = centroids_gradient,
    aes(x = NMDS1, y = NMDS2),
    color       = gradient_colors[centroids_gradient$Gradient],
    size = 8, shape = 18, show.legend = FALSE
  ) +
  ggrepel::geom_text_repel(
    data          = centroids_gradient,
    aes(x = NMDS1, y = NMDS2, label = Gradient),
    color = "black", size = 4,
    box.padding = 0.5, point.padding = 0.3,
    segment.color = "grey50", segment.size = 0.3,
    show.legend = FALSE
  ) +
  scale_color_manual(values = gradient_colors) +
  scale_fill_manual(values  = gradient_colors) +
  labs(
    title    = "NMDS of Macroinvertebrate Community Composition by Gradient",
    subtitle = paste0("Bray-Curtis dissimilarity | Stress = ",
                      round(nmds$stress, 3)),
    x = "NMDS1", y = "NMDS2",
    color = "Gradient", fill = "Gradient",
    caption = "Points = samples; diamonds = centroids; ellipses = 95% CI"
  ) +
  theme_bw() +
  theme(
    legend.position  = "right",
    panel.grid.minor = element_blank(),
    plot.title       = element_text(size = 13, face = "bold"),
    plot.subtitle    = element_text(size = 10, color = "gray40"),
    legend.text      = element_text(size = 12),
    legend.title     = element_text(size = 14)
  )

nmds_plot_gradient
```

------------------------------------------------------------------------

## Step 11: Plot NMDS — Season

```{r plot-season, fig.cap="NMDS ordination colored by collection season (Spring vs. Fall)."}

season_colors <- c(
  "Spring" = "#59A14F",
  "Fall"   = "#F28E2B"
)

nmds_plot_season <- ggplot(nmds_scores,
                           aes(x = NMDS1, y = NMDS2,
                               color = Season, fill = Season)) +
  stat_ellipse(
    aes(group = Season),
    type = "t", level = 0.95,
    geom = "polygon", alpha = 0.10, color = NA
  ) +
  stat_ellipse(
    aes(group = Season),
    type = "t", level = 0.95, linewidth = 0.6
  ) +
  geom_point(alpha = 0.6, size = 2) +
  geom_point(
    data        = centroids_season,
    aes(x = NMDS1, y = NMDS2),
    color       = season_colors[centroids_season$Season],
    size = 8, shape = 18, show.legend = FALSE
  ) +
  ggrepel::geom_text_repel(
    data          = centroids_season,
    aes(x = NMDS1, y = NMDS2, label = Season),
    color = "black", size = 4,
    box.padding = 0.5, point.padding = 0.3,
    segment.color = "grey50", segment.size = 0.3,
    show.legend = FALSE
  ) +
  scale_color_manual(values = season_colors) +
  scale_fill_manual(values  = season_colors) +
  labs(
    title    = "NMDS of Macroinvertebrate Community Composition by Season",
    subtitle = paste0("Bray-Curtis dissimilarity | Stress = ",
                      round(nmds$stress, 3)),
    x = "NMDS1", y = "NMDS2",
    color = "Season", fill = "Season",
    caption = "Points = samples; diamonds = centroids; ellipses = 95% CI"
  ) +
  theme_bw() +
  theme(
    legend.position  = "right",
    panel.grid.minor = element_blank(),
    plot.title       = element_text(size = 13, face = "bold"),
    plot.subtitle    = element_text(size = 10, color = "gray40"),
    legend.text      = element_text(size = 12),
    legend.title     = element_text(size = 14)
  )

nmds_plot_season
```

------------------------------------------------------------------------

## Step 12: Plot NMDS — Gradient × Season

```{r plot-combined, fig.cap="NMDS ordination showing the interaction of gradient type and season."}

combo_colors <- c(
  "Riffle_Spring"   = "#4E79A7",
  "Riffle_Fall"     = "#76B7B2",
  "Boatable_Spring" = "#E15759",
  "Boatable_Fall"   = "#F28E2B"
)

nmds_plot_combo <- ggplot(nmds_scores,
                          aes(x = NMDS1, y = NMDS2,
                              color = GradSeason, fill = GradSeason)) +
  stat_ellipse(
    aes(group = GradSeason),
    type = "t", level = 0.95,
    geom = "polygon", alpha = 0.10, color = NA
  ) +
  stat_ellipse(
    aes(group = GradSeason),
    type = "t", level = 0.95, linewidth = 0.6
  ) +
  geom_point(alpha = 0.6, size = 2) +
  geom_point(
    data        = centroids_combo,
    aes(x = NMDS1, y = NMDS2),
    color       = combo_colors[centroids_combo$GradSeason],
    size = 8, shape = 18, show.legend = FALSE
  ) +
  ggrepel::geom_text_repel(
    data          = centroids_combo,
    aes(x = NMDS1, y = NMDS2,
        label = gsub("_", "\n", GradSeason)),
    color = "black", size = 3.5,
    box.padding = 0.5, point.padding = 0.3,
    segment.color = "grey50", segment.size = 0.3,
    show.legend = FALSE
  ) +
  scale_color_manual(values = combo_colors,
                     labels = gsub("_", " \u2013 ", names(combo_colors))) +
  scale_fill_manual(values  = combo_colors,
                    labels  = gsub("_", " \u2013 ", names(combo_colors))) +
  labs(
    title    = "NMDS by Gradient \u00d7 Season",
    subtitle = paste0("Bray-Curtis dissimilarity | Stress = ",
                      round(nmds$stress, 3)),
    x = "NMDS1", y = "NMDS2",
    color = "Group", fill = "Group",
    caption = "Points = samples; diamonds = centroids; ellipses = 95% CI"
  ) +
  theme_bw() +
  theme(
    legend.position  = "right",
    panel.grid.minor = element_blank(),
    plot.title       = element_text(size = 13, face = "bold"),
    plot.subtitle    = element_text(size = 10, color = "gray40"),
    legend.text      = element_text(size = 11),
    legend.title     = element_text(size = 13)
  )

nmds_plot_combo
```

------------------------------------------------------------------------

## Step 13: PERMANOVA — Do Groups Differ in Community Composition?

PERMANOVA (`adonis2`) tests whether group centroids in multivariate space differ significantly. We run four models:

```{r permanova}
# Model 1: Gradient type
perm_gradient <- adonis2(
  bc_dist ~ Gradient,
  data = meta, permutations = 999, method = "bray"
)
cat("=== PERMANOVA by Gradient ===\n")
print(perm_gradient)

# Model 2: Season
perm_season <- adonis2(
  bc_dist ~ Season,
  data = meta, permutations = 999, method = "bray"
)
cat("\n=== PERMANOVA by Season ===\n")
print(perm_season)

# Model 3: Gradient + Season sequential
perm_both <- adonis2(
  bc_dist ~ Gradient + Season,
  data = meta, permutations = 999, method = "bray"
)
cat("\n=== PERMANOVA: Gradient + Season (sequential) ===\n")
print(perm_both)

# Model 4: With interaction
perm_interact <- adonis2(
  bc_dist ~ Gradient + Season + Gradient * Season,
  data = meta, permutations = 999, method = "bray"
)
cat("\n=== PERMANOVA: Gradient + Season + Interaction ===\n")
print(perm_interact)
```

> **Interpreting PERMANOVA output:** The `R²` value tells you what proportion of total community variation is explained by the grouping factor. The `Pr(>F)` is the permutation-based p-value. A significant result means group centroids are farther apart than expected by chance — but see Step 14 to check whether this reflects true centroid differences or unequal dispersion.

------------------------------------------------------------------------

## Step 14: BETADISPER — Homogeneity of Multivariate Dispersion

PERMANOVA assumes equal dispersion (variance) among groups. `betadisper` tests this assumption.

```{r betadisper}
# Dispersion by gradient
disp_gradient      <- betadisper(bc_dist, meta$Gradient)
disp_gradient_test <- permutest(disp_gradient, permutations = 999)
cat("=== Dispersion test by Gradient ===\n")
print(disp_gradient_test)

# Dispersion by season
disp_season      <- betadisper(bc_dist, meta$Season)
disp_season_test <- permutest(disp_season, permutations = 999)
cat("\n=== Dispersion test by Season ===\n")
print(disp_season_test)
```

> **Interpreting betadisper:** A non-significant result (p \> 0.05) supports the PERMANOVA assumption and means that a significant PERMANOVA reflects centroid differences. A significant betadisper result means groups also differ in within-group variability — report and discuss both.

------------------------------------------------------------------------

## Step 15: Indicator Species Analysis

Indicator species analysis (IndVal; Dufrêne & Legendre 1997) identifies taxa disproportionately associated with a particular gradient type. `multipatt` tests both specificity (taxon found mostly in one group) and fidelity (taxon found consistently within that group).

```{r indval}
indval_gradient <- multipatt(
  comm_scaled,
  meta$Gradient,
  func    = "IndVal.g",
  control = how(nperm = 999)
)

summary(indval_gradient, indvalcomp = TRUE)
```

------------------------------------------------------------------------

## Step 16: Annotate Indicators with Taxonomic Traits

The `masterTaxaGenus.csv` reference table lets you add functional feeding group (FFG) and tolerance values to indicator results — useful for ecological interpretation.

```{r taxa-traits}
indval_results <- indval_gradient$sign %>%
  tibble::rownames_to_column("FinalID") %>%
  filter(p.value < 0.05) %>%
  mutate(indicates = ifelse(s.Riffle == 1, "Riffle", "Boatable")) %>%
  left_join(
    master_taxa %>% select(FinalID, Order, Family, FFG, TolVal),
    by = "FinalID"
  ) %>%
  arrange(indicates, desc(stat))

# View top indicators with traits
indval_results %>%
  select(FinalID, indicates, stat, p.value, Order, FFG, TolVal) %>%
  head(20)
```

------------------------------------------------------------------------

## Step 17: Plot Indicator Taxa on NMDS

```{r indval-plot, fig.cap="NMDS ordination with top indicator taxa overlaid as asterisk symbols. Taxa labeled in italics are the strongest indicators of their gradient group based on IndVal.g (p < 0.05, 999 permutations)."}

top_indicators <- indval_results %>%
  group_by(indicates) %>%
  slice_max(order_by = stat, n = 5) %>%
  ungroup()

species_scores_top <- as.data.frame(scores(nmds, display = "species")) %>%
  tibble::rownames_to_column("FinalID") %>%
  inner_join(top_indicators, by = "FinalID")

cat("Indicator taxa with NMDS scores:", nrow(species_scores_top), "\n")

indicator_colors <- c("Riffle" = "#513dd1", "Boatable" = "#E15759")

nmds_indval <- nmds_plot_gradient +
  geom_point(
    data = species_scores_top,
    aes(x = NMDS1, y = NMDS2, color = indicates),
    shape = 8, size = 3, inherit.aes = FALSE
  ) +
  ggrepel::geom_text_repel(
    data        = species_scores_top,
    aes(x = NMDS1, y = NMDS2, label = FinalID, color = indicates),
    size = 3.5, fontface = "italic",
    box.padding = 0.5, point.padding = 0.3,
    segment.color = "grey60", segment.size = 0.3,
    max.overlaps = 20, inherit.aes = FALSE
  ) +
  scale_color_manual(values = indicator_colors, name = "Indicator group")

nmds_indval
```

------------------------------------------------------------------------

## Step 18: Save Outputs

```{r save-outputs, eval=FALSE}
dir.create("outputs", showWarnings = FALSE)

ggsave("outputs/NMDS_gradient.png",
       nmds_plot_gradient, width = 9, height = 6, dpi = 300)
ggsave("outputs/NMDS_season.png",
       nmds_plot_season, width = 9, height = 6, dpi = 300)
ggsave("outputs/NMDS_gradient_x_season.png",
       nmds_plot_combo, width = 10, height = 7, dpi = 300)
ggsave("outputs/NMDS_indval.png",
       nmds_indval, width = 11, height = 8, dpi = 300)

capture.output(
  list(
    PERMANOVA_gradient  = perm_gradient,
    PERMANOVA_season    = perm_season,
    PERMANOVA_both      = perm_both,
    PERMANOVA_interact  = perm_interact,
    Dispersion_gradient = disp_gradient_test,
    Dispersion_season   = disp_season_test
  ),
  file = "outputs/PERMANOVA_results.txt"
)

write.csv(indval_results, "outputs/indicator_species.csv", row.names = FALSE)

cat("All outputs saved to /outputs/\n")
```

------------------------------------------------------------------------

## Summary of Analyses

| Step | Analysis | Function | Package |
|----|----|----|----|
| 6 | log(x+1) + Wisconsin transformation | `log1p()`, `wisconsin()` | base R, `vegan` |
| 7 | Bray-Curtis dissimilarity matrix | `vegdist()` | `vegan` |
| 8 | NMDS ordination | `metaMDS()` | `vegan` |
| 13 | PERMANOVA — group centroid differences | `adonis2()` | `vegan` |
| 14 | BETADISPER — homogeneity of dispersion | `betadisper()`, `permutest()` | `vegan` |
| 15 | Indicator species analysis | `multipatt()` | `indicspecies` |
| 16 | Trait annotation | join via `masterTaxaGenus.csv` | `dplyr` |

------------------------------------------------------------------------

## References

Bray, J. R., & Curtis, J. T. (1957). An ordination of the upland forest communities of southern Wisconsin. *Ecological Monographs*, 27(4), 325–349.

De Caceres, M., & Legendre, P. (2009). Associations between species and groups of sites: indices and statistical inference. *Ecology*, 90(12), 3566–3574.

Dufrêne, M., & Legendre, P. (1997). Species assemblages and indicator species: the need for a flexible asymmetrical approach. *Ecological Monographs*, 67(3), 345–366.

McCune, B., & Grace, J. B. (2002). *Analysis of Ecological Communities*. MjM Software, Gleneden Beach, Oregon.

Oksanen, J., et al. (2022). vegan: Community Ecology Package. R package version 2.6-4. <https://CRAN.R-project.org/package=vegan>

Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis*. Springer-Verlag New York.
